Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-15 Paper_Draft2/Draft2.Rmd
finalizing draft 2

```{r race plot, echo=FALSE, fig.align='center', fig.cap= 'The race distribution for black people is 3000 and the distribution for white people is 2100. Therefore, there are more black people in the dataset than white people.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_bar_plot.png")
```

```{r age plot, echo=FALSE, fig.align='center', fig.cap= "This plot is a Kernel Density Estimate (KDE) plot. The y-axis represents the KDE of the Probability Distribution Function of the age variable in our dataset. The probability of a defendant's age being between two points on the x-axis is the total shaded area of the curve under the two points. The purple dotted line shows the average age of Black defendants and the green dotted line shows the average age of white defendants. For both groups, the majority of defendants are relatively young, but this is especially noticeable for Black defendants.", out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_age_plot.png")
```

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'Of both black and white people, there is a higher percentage of people who did not recidivate than who did. 2500 people recidivated, whereas approximately 2600 did not.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/recid_bar_plot.png")
```

```{r recid race plot,echo=FALSE, fig.align='center', fig.cap= 'When we divide the data into Black and white defendants, we can see that Black defendants recidivate more than white defendants. Black defendants recidivate are more likely than not to recidivate. However, we can also see that there are more Black defendants in the dataset in general.', out.width= "500px", out.height= "300px"}
knitr::include_graphics("../images/race_recid_plot.png")

@@

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2600 did not.', out.width= "100%"}
knitr::include_graphics("../images/recid_bar_plot_new.png")
```

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018].

helps facilitate the examination of algorithmic fairness. It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al. piece, and provides

@@

@@

                                 Result   Benefit   
                                +---------------+
Statistical Parity Difference   | -0.14 | white |
Disparate Impact Ratio          | 0.47  | white |
Average Odds Difference         | -0.44 | white |
Equal Opportunity Difference    | -0.41 | white |
                                +---------------+

>>>>

@@

```{r recid plot, echo=FALSE, fig.align='center', fig.cap= 'This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.', out.width= "100%"}
knitr::include_graphics("../images/recid_bar_plot_new.png")
```

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to "to help facilitate the transition of fairness research algorithms to use in an industrial setting and to provide a common framework for fairness researchers to share and evaluate algorithms" [@aif360-oct-2018]. It contains multiple data sets, including the COMPAS data set that accompanied the Angwin et al. piece.

@@

With our baseline model, we ran the four different group fairness metrics we chose and compared the results (pictured in Table \ref{tab:baseline metrics table}))

@@

```{r, echo = FALSE}
baseline_metrics <- tribble(~`Fairness Metric`, ~`Ideal Value`, ~ `Baseline Value`, ~`Benefited Group`,
             "Statistical Parity Difference", "0", "-0.14", "White Defendants",
             "Disparate Impact Ratio", "1", "0.47", "White Defendants",
             "Average Odds Difference", "0", "-0.44", "White Defendants",
             "Equal Opportunity Difference", "0", "-0.41", "White Defendants")
```

```{r, echo = FALSE}
knitr::kable(baseline_metrics, caption = "Results of Baseline Fairness Metric Analysis \\label{tab:baseline metrics table}")
```
